---
title: "Introduction to Optimization"
author: "Apurva Nakade"
date: "`r Sys.Date()`"
output: pdf_document
documentclass: book
link-citations: yes
github-repo: apurvanakade/2022S-Introduction-to-optimization
url: https\://apurvanakade.github.io/2022S-introduction-to-optimization/
description: Introduction to Optimization.
site: bookdown::bookdown_site
---


```{r, setup, include=FALSE}

knitr::opts_chunk$set(
  comment = NA, 
  echo = FALSE, 
  warning = FALSE, 
  cache=TRUE
)


library(ggplot2)

str_math <- function(math, inline = FALSE) {
  if (inline) {
    paste("$", str(math), "$", sep = "")
  } else {
    paste0("\\begin{align} \n", str(math), "\\end{align}\n")
  }
}

source("R/linear_program.R", local = knitr::knit_global())
source("R/linear_program_general.R", local = knitr::knit_global())
source("R/dictionaries.R", local = knitr::knit_global())
source("R/dual_linear_program.R", local = knitr::knit_global())
source("R/tableau.R", local = knitr::knit_global())

```

# Introduction

The question of optimization is the very general question of deciding when a function $f(x_1, \dots, x_n)$ attains its maximum or minimum value on a domain $D$ in $\mathbb{R^n}$. 
\begin{align}
\mbox{optimize: } && f(x_1, \dots, x_n) & \\
\mbox{subject to: } && (x_1, \dots, x_n) &\in D.
\end{align}
These kinds questions show up naturally in every quantitative field.
To analyze the question meaningfully, one needs to make some assumptions on $f$ and $D$. 

In this course, we'll analyze this question in the special case when the function $f$ is linear and the constraint set $D$ is described using linear inequalities. The study of this problem is called *Linear Programming*. Despite the simplicity of the Linear Programming setup, or perhaps because of it, LP is one of the most commonly used models for optimization problems. We'll see how to solve linear programs using the simplex method and go on to analyze the solution sets using duality theory. 

Linear programs are used for modeling real world problems when prices/costs and constraints are fixed and known beforehand.
Even when this is not the case, linear programs are often used to approximate and estimate costs/prices before moving on to more sohpisticated techniques.


::: {.example #bond-portfolio}
A bond portfolio manager has $100,000 to allocate to two different bonds; one corporate and one government bond. 
The corporate bond has a yield of 4%, a maturity of 3 years and an A rating from a rating agency that is translated into a numerical rating of 2 for computational purposes. 
In contrast, the government bond has a yield of 3%, a maturity of 6 years and rating of Aaa with the corresponding numerical rating of 1 (lower numerical ratings correspond to higher quality bonds). The portfolio manager would like to allocate her funds so that the average rating for the portfolio is no worse than Aa (numerical equivalent 1.5) and average maturity of the portfolio is at most 3.6 years. 
Any amount not invested in the two bonds will be kept in a cash account that is assumed to earn no interest for simplicity and does not contribute to the average rating or maturity computations. 
How should the manager allocate her funds between these two bonds to achieve her objective of maximizing the yield from this investment?

```{r, table-bond-portfolio, echo=FALSE, comment=NA}

df <- data.frame(
  Corporate = c("4%", 3, "A = 2", "??"),
  Government = c("3%", 6, "Aaa = 1", "??"),
  Bounds = c(" ", 3.6, 1.5, "100,000")
)
rownames(df) <- c("Yield", "Maturity", "Rating", "Fund Allocations")
df
```
:::


<details>
<summary>Solution</summary>

We can formulate the above problem as the following linear program.

\begin{equation}
  \begin{array}{rrrrrl}
  \mbox{maximize:} & 4x & + & 3y \\
  \mbox{subject to:}
    & 3x & + & 6y & \le & 3.6 \\
    & 2x & + & y & \le & 1.5 \\ 
    & x & + & y & \le & 1 \\
    & x & , & y & \ge & 0,
  \end{array}
\end{equation}

were $x$, $y$ are the percentages of corporate and government bonds respectively and the objective function when multiplied by $100,000/100 gives us the net yield. 

*We cannot subtract inequalities the same way that we can subtract equalities.* To get started, let us assume that all the money is being invested i.e. $x + y = 1$. We can then eliminate $y$ and reduce the above system to

\begin{equation}
  \begin{array}{rrrl}
  \mbox{maximize:} & x & + & 3 \\
  \mbox{subject to:}
    & 2.4 & \le & x \\
    & x & \le & 0.5.
  \end{array}
\end{equation}

This is not possible! So, it is not possible to invest all the money and still satisfy all the constraints. How does one proceed from here?
</details>