# (PART) The Simplex Method {-}

# Introduction

## What is optimization?
The question of optimization is the very general question of deciding when a function $f(x_1, \dots, x_n)$ attains its maximum or minimum value on a domain $D$ in $\mathbb{R^n}$. 
\begin{align}
\mbox{optimize: } && f(x_1, \dots, x_n) & \\
\mbox{subject to: } && (x_1, \dots, x_n) &\in D.
\end{align}
These kinds questions show up naturally in every quantitative field.
But it is almost impossible to analyze the question at this level of generality without making any further assumptions on $g$ and $D$. 

In this course, we'll analyze this question in the special case when the function $f(x)$ is linear and the constraint set $D$ is described using linear inequalities. The study of this problem is called *Linear Programming*. Despite the simplicity of the Linear Programming setup, or perhaps because of it, LP is one of the most commonly used models for real world optimization problems. We'll see how to solve linear programs using the simplex method and go on to analyze the solution sets using duality theory. 

Duality theory provides a bridge ...

## What is linear programming?
Linear programs are used for modeling real world problems when the prices and constraints are fixed and known beforehand.
These models are often used as first approximations as linear programs can be solved very efficiently and provide useful information ...
The following is an example of a *resource allocation problem*, a very common application of linear programming. 

::: {.example #resource-allocation-problem}
You run a company that makes two products (say, $P_1$ and $P_2$) using two machines (say, $M_1$ and $M_2$). Each unit of $P_1$ that is produced requires 80 minutes processing time on machine $M_1$ and 25 minutes processing time on machine $M_2$. Each unit of $P_2$ that is produced requires 20 minutes processing time on machine $M_1$ and 75 minutes processing time on machine $M_2$. Both machines are available for a maximum of 200 minutes every day. The profit per unit of $P_1$ is 25 units and the profit per unit of $P_2$ is 30 units. Company policy is to determine the production quantity of each product in such a way as to maximize the total profit given that the available resources should not be exceeded.

```{r, table-resource-allocation, echo=FALSE, comment=NA}

df <- data.frame(
  P1 = c(80, 25),
  P2 = c(20, 75),
  Upper_Bounds = c(200, 200)
)
rownames(df) <- c("M1", "M2")
df
```
:::

We can formulate the above problem as the following linear program.

```{r, latex-resource-allocation, results = 'asis', echo = FALSE}
ex_1_lp <- new_linear_program(
  obj = c(25, 30),
  A = matrix(c(80, 20, 25, 75), nrow = 2, byrow = TRUE),
  b = c(200, 200),
  variables = c('x', 'y')
)

cat(str_math(ex_1_lp))
```

Here $x$, $y$ are the units of $P_1$ and $P_2$ produced each day. One way to attempt this problem is to assume that we'll use both the machines as much as possible i.e. that the two inequalities are in fact equalities. We can see then that the common solution to the two is $(x_1, x_2) = (2, 2)$. For this solution the total profit is $110$ units.

```{r, fig-resource-allocation, fig.cap = "The feasible region is the quadrilateral formed by the overlap of the constraints."}

source("graphs/resource-allocation.R", local = knitr::knit_global())

```

::: {.exercise}
Why is this the maximum profit possible? Is it possible to increase the profit by not using either of the two machines at their full capacity?
:::


::: {.exercise #resource-allocation-problem-extra}
Consider the setup in Example \@ref(resource-allocation-problem) again. Suppose in addition, you can only store a total of 3 units of $P_1$ and $P_2$ combined each day so your production quantity should not exceed that amount. What is the new LP and the new optimal solution?
Note that the solution $(x, y) = (2, 2)$ is no longer possible. So, we cannot use both the machines to their maximum capacity.
:::

### Several definitions

More generally, a **linear program** is a problem of the following form:

\begin{equation}
  \begin{aligned}
    \mbox{optimize: } && c_1 x_1 + \dots + c_n x_n & \\
    \mbox{subject to: } 
      && a_{11} x_1 + \dots + a_{1n} x_n & \   b_1 \\
      && a_{21} x_1 + \dots + a_{2n} x_n & \lesseqqgtr b_2 \\
      && \vdots &  \\
      && a_{m1} x_1 + \dots + a_{mn} x_n & \lesseqqgtr b_m,
  \end{aligned}
  (\#eq:intro-lp)
\end{equation}
where the symbol $\lesseqqgtr$ stands for $\leq$ or $=$ or $\geq$, $a_{ij}, b_i, c_j$ are real numbers, and $x_j$ are variables.

The variables $x_1, \dots, x_n$ are called **decision variables**. The linear combination $\zeta := c_1 x_1 + \dots + c_n x_n$ is called the **objective function**. Each of the (in)equalities $a_{i1} x_1 + \dots + a_{in} x_n \lesseqqgtr b_i$ is called a **(linear) constraint**. Our goal is to either *minimize* or *maximize* the objective function subject to the constraints.

A proposal of specific values for the decision variables is called a **solution**. A solution is said to be **feasible** if it satisfies all the constraints. A feasible solution is said to be **optimal** if $\zeta$ attains the optimal value at it. Thus, to solve a linear program means to *find an optimal solution* to the problem. If a problem has no feasible solutions, then the problem is called **infeasible**. A *maximization* problem which has feasible solutions with arbitrarily large objective values is called **unbounded**. Similarly, a *minimization* problem which has feasible solutions with aribrarily small objective values is called **unbounded**.

:::{.remark}

We do not allow strict inequalities $<$ or $>$ in a linear program as linear functions do not always achieve maxima/minima on open sets. Consider the following simple example.

```{r, open-constraints, results = 'asis', echo = FALSE}
ob <- new_linear_combination(10)
c1 <- new_constraint(ob, "<", 20)
c2 <- new_constraint(new_linear_combination(5), "\\geq", 0)
cat(str_math(new_linear_program_general(ob, list(c1, c2))))
```

On the feasible set $[0, 2)$, the function $\zeta(x_1) = 10x_1$ never attains absolute maxima.
Changing the inequality $<$ to $\leq$ gives us an optimal feasible solution $x_1 = 2$ and $\zeta = 20$. When all the (in)equalities are either $\le$, $=$, or $\ge$, the set of feasible solutions is closed. If, in addition, the LP is bounded then the set of feasible solutions is compact. On compact sets a continuous (in particular, linear) function, always attains a maxima and a minima by a generalization of the [extreme value theorem](https://en.wikipedia.org/wiki/Extreme_value_theorem) for [higher dimensions](https://math.stackexchange.com/a/881569).

:::


::: {.remark}

"Unboundedness" means different things depending on whether we're maximizing or minimizing the optimization function. For example, the following linear program
...
is not *unbounded* but linear program
...
is *unbounded*.

:::
